ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-14 paper/paper.tex
some working on cutting down pages

Counterfactual Explanations provide a powerful, flexible and intuitive way to not only explain black-box models but also enable affected individuals to challenge them through the means of Algorithmic Recourse. Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}. Intuitively speaking, we generate explanations in this context by asking simple what-if questions of the following nature: `Our credit risk model currently predicts that this individual's credit profile is too risky to offer them a loan. What if they reduced their monthly expenditures by 10\%? Will our model then predict that the individual is credit-worthy'?

This is typically implemented by defining a target outcome $\mathbf{y}^+ \in \mathcal{Y}$ for some individual $\mathbf{x} \in \mathcal{X}=\mathbb{R}^D$ described by $D$ attributes, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(\mathbf{x})\ne \mathbf{y}^+$. Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$. Since Counterfactual Explanations (CE) work directly with the black-box model, valid counterfactuals always have full local fidelity by construction~\citep{mothilal2020explaining}. Fidelity is defined as the degree to which explanations approximate the predictions of the black-box model. This is arguably one of the most important evaluation metrics for model explanations, since any explanation that explains a prediction not actually made by the model is useless~\citep{molnar2020interpretable}.

In situations where full fidelity is a requirement, CE therefore offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models. But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model. That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks which are typically underspecified by the available data~\citep{wilson2020case}.

In the context of CE, the idea that no two explanations are the same arises almost naturally. A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are deemed most appropriate based on a myriad of desiderata such as sparsity, actionability and plausibility. In this work, we draw closer attention to the insufficiency of model fidelity as an evaluation metric for the faithfulness of counterfactual explanations. Our key contributions are as follows: firstly, we introduce a new notion of faithfulness that is suitable for counterfactuals and propose a novel evaluation metric that draws inspiration from recent advances in Energy-Based Modelling (EBM); secondly, we a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCo) that explicitly address the need for faithfulness; finally, we provide illustrative examples and extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour without sacrificing existing desidarata like plausibility and sparsity.

@@

In this section, we provide some background on Counterfactual Explanations and our motivation for this work. To start, we briefly introduce the methodology underlying most state-of-the-art (SOTA) counterfactual generators.

@@

While Counterfactual Explanations can be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems. Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes. Then most SOTA counterfactual generators rely on gradient descent to optimize different flavours of the following counterfactual search objective:

@@

Here $\text{yloss}$ denotes the primary loss function already introduced above and $\text{cost}$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization. Equation~\ref{eq:general} restates the baseline approach to gradient-based counterfactual search proposed by~\citet{wachter2017counterfactual} in general form where $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states~\citep{altmeyer2023endogenous}. This is to explicitly account for the multiplicity of explanations and the fact that we may choose to generate multiple counterfactuals and traverse a latent encoding $\mathcal{Z}$ of the feature space $\mathcal{X}$ where we denote $f^{-1}: \mathcal{X} \mapsto \mathcal{Z}$. Encodings may involve simple feature transformations or more advanced techniques involving generative models, as we will discuss further below. The baseline approach, which we will simply refer to as \textbf{Wachter}~\citep{wachter2017counterfactual}, searches a single counterfactual directly in the feature space and penalises its distance between the original factual.

@@

Arguably, the plausibility of counterfactuals has been among the primary concerns and some have focused explicitly on this goal. \citet{joshi2019realistic}, for example, were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the data generating process (DGP) of $\mathbf{x}\sim\mathcal{X}$. To learn the latent embedding, they introduce a surrogate model. In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $\mathbf{x}^* \leftarrow \mathcal{G}(\mathbf{z})$ where $\mathcal{G}$ denotes the decoder part of the VAE. Provided the surrogate model is well-trained, their proposed approach ---REVISE--- can yield compelling counterfactual explanations like the one in the centre panel of Figure~\ref{fig:vae}.

@@

\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution. Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead. Their proposed methodology solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all. They demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence. Provided the model is well-specified, these counterfactuals are plausible. Unfortunately, this idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.

@@

Above we explained that since Counterfactual Explanations work directly with the Black Box model, the fidelity of explanations as we defined it earlier is not a concern. This may explain why research has primarily focused on other desiderata, most notably plausibility (Definition~\ref{def:plausible}). Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the underlying data'? We posit a related, slightly more nuanced question: `Is this counterfactual consistent with what the model has learned about the underlying data'? We will argue that fidelity is not a sufficient evaluation metric to answer this question and propose a novel way to assess if Counterfactual Explanations conform with model behaviour.

The word \textit{fidelity} stems from the Latin word `fidelis', which means `faithful, loyal, trustworthy'~\citep{mw2023fidelity}. As we explained in Section~\ref{background}, model explanations are generally considered faithful if their corresponding predictions coincide with the predictions made by the model itself. Since this definition of faithfulness is not useful in the context of Counterfactual Explanations, we propose an adapted version:

@@

In words, conformal counterfactuals conform with what the predictive model has learned about the input data $\mathbf{x}$. Since this definition works with distributional properties, it explicitly accounts for the multiplicity of explanations we discussed earlier. To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$. This is very much at the core of our proposed methodological framework, which reconciles the notions of plausibility and model faithfulness and which we will introduce next.

@@

Recent work by~\citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'. The authors show that we can draw samples from the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using Stochastic Gradient Langevin Dynamics (SGLD). The authors use this insight to train classifiers jointly for the discriminative task using standard cross-entropy and the generative task using SGLD. They demonstrate empirically that among other things this improves predictive uncertainty quantification for discriminative models. Our findings in this work suggest that Joint Energy Models (JEM) also tend to yield more plausible Counterfactual Explanations. Based on the definition of plausible counterfactuals (Definition~\ref{def:plausible}) this is not surprising.

Crucially for our purpose, one can apply their proposed sampling strategy during inference to essentially any standard discriminative model. Even models that are not explicitly trained for the joint objective learn about the distribution of inputs $X$ by learning to make conditional predictions about the output $y$. We can leverage this observation to quantify the generative property of the Black Box model itself. In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can sample from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ using SGLD as follows,

@@

where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed. The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the energy function where we use $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)=-M_{\theta}(\mathbf{x}_j)[\mathbf{y}^+]$, that is the negative logit corresponding to the target class label $\mathbf{y}^+$. Generating multiple samples in this manner yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that we use in our search for plausible counterfactuals, as discussed in more detail below. Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.

@@

where $\hat{\mathbf{x}}_{\theta}$ denotes samples generated using SGLD (Equation~\ref{eq:sgld}) and $\text{dist}(\cdot)$ is a generic term for a distance metric. Our default choice for $\text{dist}(\cdot)$ is the L1 Norm, or Manhattan distance, since it induces sparsity.

@@

\begin{minipage}[c]{0.40\textwidth}
  \centering
  \includegraphics[width=\textwidth]{../artifacts/results/images/poc.png}
  \captionof{figure}{[PLACEHOLDER; may swap for:] Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
\end{minipage}  
\hfill
\begin{minipage}[c]{0.50\textwidth}
\captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
\begin{algorithmic}[1]
  \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
  \Ensure $\mathbf{x}^\prime$
  \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^+$ using SGLD (Equation~\ref{eq:sgld})
  \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
  \State Initialize $t \gets 0$
  \While{\textit{not converged} or $t < T$}
  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$
  \State $t \gets t+1$
  \EndWhile
  \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$
\end{algorithmic}
\end{minipage}

@@

\begin{minipage}[c]{\textwidth}
  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
  \captionof{figure}{Original image (left) and ECCCos for turning a 9 (nine) into a 7 (seven) for different black-box models from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
\end{minipage}

@@

In this section, we present our empirical analysis and findings. Our goal is to shed line on the following questions:

@@

In the following, we first briefly describe our evaluation framework and experimental setup, before presenting and discussing our results.

@@

Above we have defined plausibility (Definition~\ref{def:plausible}) and faithfulness (Definition~\ref{def:faithful}) for Counterfactual Explanations. These are the main criteria we use to evaluate counterfactuals in this study. In order to quantify the plausibility of counterfactuals we use a slightly adapted version of the implausibility metric proposed in~\citet{guidotti2022counterfactual}. Formally, for a single counterfactual, we define implausibility as follows,

@@

Specifically, we form this subset based on the $n_E$ generated samples associated with the lowest energy.

@@

We use both synthetic and real-world datasets from different domains, all of which are publically available and commonly used to train and benchmark classification algorithms. The synthetic datasets include: a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data. As for real-world data, we follow~\citet{schut2021generating} and use the \textbf{MNIST}~\citep{lecun1998mnist} image dataset. It is composed of 60000 images of 28x28 pixels each showing handwritten digits from 0 to 9 such as the examples shown above. From the social sciences domain, we include Give Me Some Credit (\textbf{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}. It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether or not retail borrowers experience financial distress.

As with the example in Section~\ref{meth}, we use simple neural networks (\textbf{MLP}), ensembles of neural networks (\textbf{MLP Ensemble}), Joint Energy Models (\textbf{JEM}) and ensembles of JEMs (\textbf{JEM Ensemble}) to model our real-world datasets. For the synthetic datasets, we found that the use of ensembles was not necessary.

To account for stochasticity, we generate multiple counterfactuals for each possible target class, generator, model and dataset. Specifically, we randomly sample $n^{-}$ times from the subset of individuals for which the given model predicts the non-target class $\mathbf{y}^{-}$ given the current target. We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for GMSC and $n^{-}=5$ for MNIST. Note that in the latter case, we still end up generating disproportionately more counterfactuals because all 10 digits appear both as factuals and targets for each other. Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.

@@

@@

Starting with the high-level results for our Linearly Separable data, we find that ECCCo produces the most faithful counterfactuals for both black-box models. This is not surprising, since ECCCo directly enforces faithfulness through regularization. Crucially though, ECCCo also produces the most plausible counterfactuals for the Joint Energy Model, which was explicitly trained to learn plausible representations of the input data. This high-level pattern is broadly consistent across all datasets and supportive of our narrative, so it is worth highlighting it here: ECCCos consistently achieve high faithfulness, which---subject to the quality of the black-box model itself---coincides with high plausibility.

Zooming in on the granular details for the Linearly Separable data, note that the list of generators in Table~\ref{tab:results-synthetic} includes `ECCCo (no CP)' and `ECCCo (no EBM)' in addition to `ECCCo' and our benchmark generators. These have been added to gain some sense of the degree to which the two components underlying ECCCo---namely energy-based modelling (EBM) and conformal prediction (CP)---drive the results. Specifically, `ECCCo (no CP)' involves no set size penalty ($\lambda_3=0$ in Equation~\ref{eq:eccco}), while `ECCCo (no EBM)' does not penalise the distance to samples generated through SGLD ($\lambda_2=0$ in Equation~\ref{eq:eccco}). The corresponding results indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM) in our search for counterfactuals. Conformal Prediction alone only leads to marginally improved faithfulness and plausibility relative to the benchmark generators for our JEM.

As a final key observation for the Linearly Separable data we note that for the MLP, increased faithfulness comes at the cost of reduced plausibility. Specifically, this means that counterfactuals generated through ECCCo end up further away from individuals in the target class than those produced by our benchmark generators.

The findings for the Moons dataset are broadly in line with the findings so far: for the JEM, ECCCo yields significantly more faithful and plausible counterfactuals than all other generators. For the MLP, faithfulness is maintained but counterfactuals are not plausible. By comparison, REVISE yields fairly plausible counterfactuals in both cases, but it does so at the cost of faithfulness. We also observe that the best results for ECCCo are achieved when using both penalties. Once again though, the generative component (EBM) has a stronger impact on the positive results for the JEM. a

@@

\textit{NOTE: visual inspection of the results shows that ECCCo still does exactly what it's supposed to do. I think the outcome here demonstrates the limitations of the implausibility metric: computing average distances to randomly drawn samples from the target class does not really make perfect sense when samples are distributed as a circle. But this story may be too convoluted to add here?}

@@

For the tabular credit dataset (GMSC) we have struggled to get good generative and discriminative performance for our JEMs. Consequently, it is not surprising to find that ECCCo never achieves state-of-the-art plausibility, although it does improve outcomes compared to the baseline (Wachter). With respect to faithfulness, ECCCo once again consistently outperforms all other counterfactual generators.

@@

To conclude this section, we summarize our findings with reference to the opening questions. Concerning the feasibility of our proposed methodology (Research Question~\ref{rq:feasibility}), our findings clearly demonstrate that it is indeed possible to generate plausible counterfactuals without the need for surrogate models. A related important finding is that ECCCo never sacrifices faithfulness for plausibility: any plausible ECCCo also faithfully describes model behaviour. This mitigates the risk of generating plausible explanations for models that are, in fact, highly susceptible to implausible counterfactuals as well. Our findings here indicate that ECCCo achieves this result primarily by leveraging the model's generative property. We think that further work is needed, however, to definitively answer Research Question~\ref{rq:drivers}, on which we elaborate in the following section.

>>>>

Counterfactual Explanations (CE) provide a powerful, flexible and intuitive way to not only explain black-box models but also help affected individuals through the means of Algorithmic Recourse. Instead of opening the Black Box, CE works under the premise of strategically perturbing model inputs to understand model behaviour~\citep{wachter2017counterfactual}. Intuitively speaking, we generate explanations in this context by asking what-if questions of the following nature: `Our credit risk model currently predicts that this individual is not credit-worthy. What if they reduced their monthly expenditures by 10\%?'

This is typically implemented by defining a target outcome $\mathbf{y}^+ \in \mathcal{Y}$ for some individual $\mathbf{x} \in \mathcal{X}=\mathbb{R}^D$ described by $D$ attributes, for which the model $M_{\theta}:\mathcal{X}\mapsto\mathcal{Y}$ initially predicts a different outcome: $M_{\theta}(\mathbf{x})\ne \mathbf{y}^+$. Counterfactuals are then searched by minimizing a loss function that compares the predicted model output to the target outcome: $\text{yloss}(M_{\theta}(\mathbf{x}),\mathbf{y}^+)$. Since Counterfactual Explanations work directly with the black-box model, valid counterfactuals always have full local fidelity by construction where fidelity is defined as the degree to which explanations approximate the predictions of a black-box model~\citep{mothilal2020explaining,molnar2020interpretable}.

In situations where full fidelity is a requirement, CE offers a more appropriate solution to Explainable Artificial Intelligence (XAI) than other popular approaches like LIME~\citep{ribeiro2016why} and SHAP~\citep{lundberg2017unified}, which involve local surrogate models. But even full fidelity is not a sufficient condition for ensuring that an explanation faithfully describes the behaviour of a model. That is because multiple very distinct explanations can all lead to the same model prediction, especially when dealing with heavily parameterized models like deep neural networks which are typically underspecified by the data~\citep{wilson2020case}.

In the context of CE, the idea that no two explanations are the same arises almost naturally. A key focus in the literature has therefore been to identify those explanations and algorithmic recourses that are most appropriate based on a myriad of desiderata such as sparsity, actionability and plausibility. In this work, we draw closer attention to model faithfulness as a desideratum for counterfactuals. Our key contributions are as follows: firstly, we propose a definition of faithfulness that is suitable for counterfactuals; secondly, we introduce a novel algorithmic approach for generating Energy-Constrained Conformal Counterfactuals (ECCCos) that explicitly addresses the need for faithfulness; finally, we provide extensive empirical evidence demonstrating that ECCCos faithfully explain model behaviour without sacrificing plausibility.

@@

In this section, we provide some background on CE and our motivation for this work. To start, we briefly introduce the methodology underlying most state-of-the-art (SOTA) counterfactual generators.

@@

While Counterfactual Explanations can be generated for arbitrary regression models~\citep{spooner2021counterfactual}, existing work has primarily focused on classification problems. Let $\mathcal{Y}=(0,1)^K$ denote the one-hot-encoded output domain with $K$ classes. Then most counterfactual generators rely on gradient descent to optimize different flavours of the following counterfactual search objective:

@@

Here $\text{yloss}$ denotes the primary loss function already introduced above and $\text{cost}$ is either a single penalty or a collection of penalties that are used to impose constraints through regularization. Equation~\ref{eq:general} restates the baseline approach to gradient-based counterfactual search proposed by~\citet{wachter2017counterfactual} in general form where $\mathbf{Z}^\prime=\{ \mathbf{z}_l\}_L$ denotes an $L$-dimensional array of counterfactual states~\citep{altmeyer2023endogenous}. This is to explicitly account for the multiplicity of explanations and the fact that we may choose to generate multiple counterfactuals and traverse a latent encoding $\mathcal{Z}$ of the feature space $\mathcal{X}$ where we denote $f^{-1}: \mathcal{X} \mapsto \mathcal{Z}$. Encodings may involve simple feature transformations or more advanced techniques involving generative models, as we will discuss further below. The baseline approach, which we will simply refer to as \textbf{Wachter}~\citep{wachter2017counterfactual}, searches a single counterfactual directly in the feature space and penalises its distance to the original factual.

@@

Arguably, the plausibility of counterfactuals has been among the primary concerns and some have focused explicitly on this goal. \citet{joshi2019realistic}, for example, were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ (Equation~\ref{eq:general}) that implicitly codifies the data generating process (DGP) of $\mathbf{x}\sim\mathcal{X}$. To learn the latent embedding, they introduce a surrogate model. In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $\mathbf{x}^* \leftarrow \mathcal{G}(\mathbf{z})$ where $\mathcal{G}$ denotes the decoder part of the VAE. Provided the surrogate model is well-trained, their proposed approach ---\textbf{REVISE}--- can yield plausible explanations like the one in the centre panel of Figure~\ref{fig:vae}.

@@

\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution. Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead. Their proposed methodology solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all. They demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^+$ with high confidence. Provided the model is well-specified, these counterfactuals are plausible. This idea hinges on the assumption that the black-box model provides well-calibrated predictive uncertainty estimates.

@@

Above we explained that since Counterfactual Explanations work directly with the Black Box model, the fidelity of explanations as we defined it earlier is not a concern. This may explain why research has primarily focused on other desiderata, most notably plausibility (Definition~\ref{def:plausible}). Enquiring about the plausibility of a counterfactual essentially boils down to the following question: `Is this counterfactual consistent with the data'? We ask a related question: `Is this counterfactual consistent with what the model has learned about the data'? To answer this question and propose a novel way to assess if explanations conform with model behaviour.

The word \textit{fidelity} stems from the Latin word `fidelis', which means `faithful, loyal, trustworthy'~\citep{mw2023fidelity}. As we explained in Section~\ref{background}, model explanations are generally considered faithful if their corresponding predictions coincide with the predictions made by the model itself. Since this definition of faithfulness is not useful in the context of CE, we propose the following definition:

@@

To assess counterfactuals with respect to Definition~\ref{def:faithful}, we need to be able to quantify the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$. This is very much at the core of our proposed methodological framework, which we will introduce next.

@@

Recent work by~\citet{grathwohl2020your} on Energy-Based Models (EBM) has shown that there is a `generative model hidden within every standard discriminative model'. They use Stochastic Gradient Langevin Dynamics (SGLD) to estimate the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using and leverage this to train classifiers jointly for the discriminative and generative task.

Crucially for our purpose, SGLD can be applied during inference to essentially any standard discriminative model. Even models that are not explicitly trained for the joint objective learn about the distribution of inputs $X$ by learning to make conditional predictions about the output $y$. We can leverage this observation to quantify the generative property of models. In particular, note that if we fix $\mathbf{y}$ to our target value $\mathbf{y}^+$, we can sample from $p_{\theta}(\mathbf{x}|\mathbf{y}^+)$ using SGLD as follows,

@@

where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon$ is typically polynomially decayed. The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$. Generating multiple samples in this manner yields an empirical distribution $\hat{\mathbf{X}}_{\theta,\mathbf{y}^+}$ that we use in our search for plausible counterfactuals, as discussed in more detail below. Appendix~\ref{app-jem} provides additional implementation details for any tasks related to energy-based modelling.

@@

where $\hat{\mathbf{x}}_{\theta}$ denotes samples generated using SGLD (Equation~\ref{eq:sgld}) and $\text{dist}(\cdot)$ is a generic term for a distance metric. Our default choice for $\text{dist}(\cdot)$ is the L1 Norm since it induces sparsity.

@@

\begin{minipage}[c]{0.45\textwidth}
  \centering
  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/poc.png}
  \captionof{figure}{[PLACEHOLDER; may swap for:] Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
\end{minipage}  
\hfill
\begin{minipage}[c]{0.45\textwidth}
\captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
\begin{algorithmic}[1]
  \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
  \Ensure $\mathbf{x}^\prime$
  \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^+$ using SGLD (Equation~\ref{eq:sgld})
  \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
  \State Initialize $t \gets 0$
  \While{\textit{not converged} or $t < T$}
  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$
  \State $t \gets t+1$
  \EndWhile
  \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$
\end{algorithmic}
\end{minipage}

@@

\begin{minipage}[c]{\textwidth}
  \centering
  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_eccco.png}
  \captionof{figure}{Original image (left) and ECCCos for turning a 9 (nine) into a 7 (seven) for different black-box models from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
\end{minipage}

@@

Our goal in this section is to shed light on the following research questions:

@@

We first briefly describe our experimental setup, before presenting our main results.

@@

Above we have defined plausibility (Definition~\ref{def:plausible}) and faithfulness (Definition~\ref{def:faithful}) for Counterfactual Explanations. These are the main criteria we use to evaluate counterfactuals in this study. In order to quantify the plausibility of counterfactuals we use a slightly adapted version of the implausibility metric proposed in~\citet{guidotti2022counterfactual}. Formally, we define implausibility as follows,

@@

Specifically, we form this subset based on the $n_E$ generated samples with the lowest energy.

@@

We use both synthetic and real-world datasets from different domains, all of which are publically available and commonly used to train and benchmark classification algorithms. The synthetic datasets include: a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data. As for real-world data, we follow~\citet{schut2021generating} and use the \textbf{MNIST}~\citep{lecun1998mnist} dataset containing images of handwritten digits such as the examples shown above. From the social sciences domain, we include Give Me Some Credit (\textbf{GMSC})~\citep{kaggle2011give}: a tabular dataset that has been studied extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}. It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether or not retail borrowers experience financial distress.

As with the example in Section~\ref{meth}, we use simple neural networks (\textbf{MLP}) and Joint Energy Models (\textbf{JEM}). For the more complex real-world datasets we also use ensembling in each case. To account for stochasticity, we generate multiple counterfactuals for each possible target class, generator, model and dataset. Specifically, we randomly sample $n^{-}$ times from the subset of individuals for which the given model predicts the non-target class $\mathbf{y}^{-}$ given the current target. We set $n^{-}=25$ for all of our synthetic datasets, $n^{-}=10$ for GMSC and $n^{-}=5$ for MNIST. Full details concerning our parameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.

@@

Starting with the high-level results for our Linearly Separable data, we find that ECCCo produces the most faithful counterfactuals for both black-box models. This is not surprising, since ECCCo directly enforces faithfulness through regularization. Crucially though, ECCCo also produces the most plausible counterfactuals for the Joint Energy Model, which was explicitly trained to learn plausible representations of the input data. This high-level pattern is broadly consistent across all datasets and supportive of our narrative, so it is worth highlighting: ECCCos consistently achieve high faithfulness, which---subject to the quality of the model itself---coincides with high plausibility.

Zooming in on the granular details for the Linearly Separable data, note that the list of generators in Table~\ref{tab:results-synthetic} includes `ECCCo (no CP)' and `ECCCo (no EBM)' in addition to `ECCCo' and our benchmark generators. These have been added to gain some sense of the degree to which the two components underlying ECCCo---namely energy-based modelling (EBM) and conformal prediction (CP)---drive the results. Specifically, `ECCCo (no CP)' involves no set size penalty ($\lambda_3=0$ in Equation~\ref{eq:eccco}), while `ECCCo (no EBM)' does not penalise the distance to samples generated through SGLD ($\lambda_2=0$ in Equation~\ref{eq:eccco}). The corresponding results indicate that the positive results are dominated by the effect of quantifying and leveraging the model's generative property (EBM) in our search for counterfactuals. Conformal Prediction alone only leads to marginally improved faithfulness and plausibility relative to the benchmark generators for our JEM. As a final observation for the Linearly Separable data we note that for the MLP, increased faithfulness comes at the cost of reduced plausibility. Specifically, this means that counterfactuals generated through ECCCo end up further away from individuals in the target class than those produced by our benchmark generators.

The findings for the Moons dataset are broadly in line with the findings so far: for the JEM, ECCCo yields significantly more faithful and plausible counterfactuals than all other generators. For the MLP, faithfulness is maintained but counterfactuals are not plausible. By comparison, REVISE yields fairly plausible counterfactuals in both cases, but it does so at the cost of faithfulness. We also observe that the best results for ECCCo are achieved when using both penalties. Once again though, the generative component (EBM) has a stronger impact on the positive results for the JEM.

@@

@@

@@

For the tabular credit dataset (GMSC) we have struggled to get good generative and discriminative performance for our JEMs. Consequently, it is not surprising to find that ECCCo never achieves state-of-the-art plausibility, although it does improve outcomes compared to the baseline (Wachter). Concerning faithfulness, ECCCo once again consistently outperforms all other generators.

To conclude this section, we summarize our findings with reference to the opening questions. Concerning the feasibility of our proposed methodology (Research Question~\ref{rq:feasibility}), our findings demonstrate that it is indeed possible to generate plausible counterfactuals without the need for surrogate models. A related important finding is that ECCCo never sacrifices faithfulness for plausibility: any plausible ECCCo also faithfully describes model behaviour. This mitigates the risk of generating plausible explanations for models that are, in fact, highly susceptible to implausible counterfactuals as well. Our findings here indicate that ECCCo achieves this result primarily by leveraging the model's generative property. We think that further work is needed, however, to definitively answer Research Question~\ref{rq:drivers}, on which we elaborate in the following section.

@@
